---
title: "Getting started"
output:
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 9, fig.height = 6, fig.align = "center",
  warning = FALSE, message = FALSE, dev = "png", dpi = 150
)
```

```{r load}
library(censusindia)
library(dplyr)
library(ggplot2)
```

## What is included in censusindia?

Census data for every decennial census from 1901 to 2011. Some years also carry
the Primary Census Abstract, which adds households, literacy, Scheduled Caste
and Tribe counts and work status to the population totals. The 2011 abstract
goes down to village and ward.



```{r inventory}
span <- function(v) {
  v <- sort(unique(v[!is.na(v)]))
  if (!length(v)) {
    return(NA_character_)
  }
  if (length(v) <= 3) paste(v, collapse = ", ") else sprintf("%d-%d (%d)", min(v), max(v), length(v))
}

grain <- function(d) {
  for (col in c("level", "geography")) {
    if (col %in% names(d)) {
      return(paste(sort(unique(d[[col]])), collapse = "/"))
    }
  }
  if (any(c("district", "district_name") %in% names(d))) "district" else NA_character_
}

vapply(sort(utils::data(package = "censusindia")$results[, "Item"]), function(n) {
  d <- get(n, envir = asNamespace("censusindia"))
  yr <- if ("year" %in% names(d)) span(d$year) else stringr::str_extract(n, "\\d{4}")
  c(
    years = yr %||% "-", grain = grain(d) %||% "-",
    rows = format(nrow(d), big.mark = ","), cols = as.character(ncol(d))
  )
}, character(4)) |>
  t() |>
  knitr::kable()
```

Population is available across all twelve censuses, 1901 to 2011. 
The detailed abstracts exist for five years only: 1961 (literacy), 1971, 1981, 2001 and
2011. 

## Attaching maps

Any indicator can go straight onto the boundaries of its census year:

```{r first-map, fig.height = 7}
get_census(2011, "district") |>
  attach_geometry(2011, geography = "district") |>
  plot_map("population",
    title = "District population, 2011",
    legend_title = "Persons", palette = "blues", trans = "log10"
  )
```

## Census years are not interchangeable

Each year has its own source table, and they carry different columns. 1961 is
a literacy table, district level only. 1971 and 1981 split by sector. 2011
reaches subdistricts.

```{r geographies}
list_census_geographies() |> count(geography)
```

If you ask for a year and geography that do not exist together, you will get an error
rather than an empty frame:

```{r not-available, error = TRUE}
get_census(1961, "state")
```

For 1971 and 1981, `sector` defaults to `"total"`. Rural and urban are there too:
```{r sector}
c(
  total = sum(get_census(1981, "state")$population),
  urban = sum(get_census(1981, "state", sector = "urban")$population)
)
```

## Finding a variable

Column names differ between years, so `variables =` takes canonical names and
maps them onto whatever that year calls them. `list_census_variables()` prints
all 23.

```{r search}
search_census_variables("literacy")

get_census(2011, "state", variables = c("population", "literacy_rate")) |> head(3)
```

Filter states by name or two-letter code: 

```{r states}
get_census(1971, "district", state = "MH") |>
  select(name, population) |>
  head(3)

list_states(region = "South")$state_name
```



For every year, `get_census()` returns the population time series.
The detailed abstracts are separate tables:

```{r pca-vs-timeseries}
c(
  get_census_2001 = nrow(get_census(2001, "district")),
  census_2001_pca = nrow(census_2001_pca),
  get_census_2011 = nrow(get_census(2011, "district")),
  census_2011_pca = nrow(census_2011_pca)
)
```

Literacy, SC/ST and worker counts live in the abstracts, for 2001 and 2011:

```{r pca}
census_2001_pca |>
  summarise(
    districts = n(),
    literate_pct = 100 * sum(literate_total) / sum(population_total),
    worker_pct = 100 * sum(workers_total) / sum(population_total)
  )
```

`census_2001_pca_full` is the same 593 districts with the full published
abstract: nineteen indicators, most split by sex, each district given three
times for total, rural and urban.

```{r pca-indicators}
ind <- setdiff(
  names(census_2001_pca_full),
  c(
    "year", "geography", "state_code", "state_name",
    "state_name_harmonized", "district_code", "name", "sector"
  )
)

tibble::tibble(
  indicator = sub("_(total|male|female)$", "", ind),
  part = sub("^.*_", "", ind)
) |>
  group_by(indicator) |>
  summarise(split_by = paste(sort(unique(part)), collapse = ", "), .groups = "drop") |>
  knitr::kable()
```

Which lets you split rural from urban:

```{r pca-demo}
census_2001_pca_full |>
  filter(sector != "total") |>
  group_by(sector) |>
  summarise(
    cultivators = 100 * sum(main_cultivators_total) / sum(main_workers_total),
    agri_labour = 100 * sum(main_agri_labour_total) / sum(main_workers_total),
    hh_industry = 100 * sum(main_hh_industry_total) / sum(main_workers_total),
    other = 100 * sum(main_other_total) / sum(main_workers_total),
    .groups = "drop"
  )
```

1971 and 1981 have their own abstracts in `census_1971` and `census_1981`;
1961 has literacy only.

## Attaching geometry

Districts split and get renamed between censuses. `attach_geometry()`allows
you to visulaise the census specific boundaries:

```{r boundaries, fig.height = 6.5, fig.width = 9}
YEARS <- c(1941, 1951, 1961, 1971, 1981, 1991, 2001, 2011)

lapply(YEARS, function(y) {
  get_census_boundaries(y, "state") |>
    sf::st_geometry() |>
    sf::st_as_sf() |>
    transform(year = y)
}) |>
  do.call(what = rbind) |>
  ggplot() +
  geom_sf(fill = "grey95", colour = "grey40", linewidth = 0.15) +
  facet_wrap(~year, nrow = 2) +
  theme_void(base_size = 10) +
  theme(strip.text = element_text(face = "bold"))
```


```{r district-vintage}
rate <- function(d, y) {
  g <- suppressWarnings(attach_geometry(d, year = y, geography = "district"))
  sprintf("%d/%d", sum(!sf::st_is_empty(g$geometry)), nrow(d))
}

c(
  `1971 abstract + 1971` = rate(filter(census_1971, geography == "district"), 1971),
  `time series 1971 + 1971` = rate(
    filter(census_population_time_series, geography == "district", year == 1971), 1971
  )
)
```

District maps need the abstract for that year. 

Boundary files are parsed once and held for the session. `clear_geometry_cache()`
releases them, `compare_maps()` puts several years on one colour scale, and
`get_palette()` returns the ramps `plot_map()` draws with.

```{r palette}
get_palette("blue_red")
clear_geometry_cache()
```



## Population Projections

`get_population()` reads the [MoHFW projections](https://main.mohfw.gov.in/sites/default/files/Population%20Projection%20Report%202011-2036%20-%20upload_compressed_0.pdf),
which run past the last census. The state table is annual, 2011 to 2036, with
males and females for 36 states plus India and Ladakh.

```{r projections-shape}
proj <- get_population(geography = "state")

c(
  entries = dplyr::n_distinct(proj$state_name_harmonized),
  years = dplyr::n_distinct(proj$year),
  first = min(proj$year), last = max(proj$year)
)
```


```{r projections-trend, fig.height = 4}
proj |>
  filter(state_name_harmonized == "India") |>
  arrange(year) |>
  mutate(added = population - lag(population)) |>
  filter(!is.na(added)) |>
  ggplot(aes(year, added / 1e6)) +
  geom_col(fill = "#08519C", width = 0.7) +
  labs(x = NULL, y = "Millions added that year") +
  theme_minimal()
```

District projections exist on two geographies: `census2011` has the 640 Census districts at five-year steps.
`lgd` has 785 present-day districts annually (note that there are not scaled and hence should not be used downstream without reclaibrating).

```{r projections-district}
c(
  census2011 = nrow(get_population(2031, "district")),
  lgd = nrow(get_population(2031, "district", boundary = "lgd"))
)
```

## Going below district level

2011 also ships the abstract at the village/ward level:

```{r detail-use}
census_2011_demographics |>
  filter(level == "district", sector == "total") |>
  summarise(districts = n(), population = sum(population_total))
```

Worker classification is available in the `census_2011_workers` table and
`census_2011_marginal_detail` provides the marginal-worker duration split.
Tribal and language information is also available:

```{r communities}
census_2011_tribes |>
  slice_max(population, n = 3) |>
  select(state_name_harmonized, district_name, tribe_name, population)

census_2011_linguistic_diversity |>
  filter(coverage >= 0.9) |>
  slice_max(effective_languages, n = 3) |>
  select(district_name, n_languages, effective_languages, coverage)
```

`coverage` is the share of the district population C-16 gives a named mother
tongue. It drops to 0.04 in parts of Nagaland, where the unscheduled languages
fall into unnamed residual categories, so an unfiltered ranking puts those
districts on top.

