---
title: "Mapping linguistic diversity"
output:
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Mapping linguistic diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 10,
  warning = FALSE,
  message = FALSE,
  dev = "png",
  dpi = 150
)
```

```{r load-packages}
library(censusindia)
library(dplyr)
library(ggplot2)
library(patchwork)
```

```{r load-data}
data(census_2011_mother_tongue)

census_2011_mother_tongue |>
  filter(district_code == "000") |>
  group_by(language_level) |>
  summarise(n_languages = n_distinct(language_name))
```

Both measures below are computed from each language's population share $p_i$
in a district, but they weight rare languages differently:

- **Shannon entropy**, $S = -\sum_i p_i \log(p_i)$. We report it as the
  *effective number of languages*, $2^S$: Hill's (1973) "numbers
  equivalent," the count of equally-sized languages that would produce the
  same $S$. It is more sensitive to a long tail of small languages than a
  raw language count (Grin & Fürst, 2022).
- **Greenberg's Diversity Index**, $G = 1 - \sum_i p_i^2$ (Greenberg, 1956).
  It is the probability that two speakers picked at random from the
  district speak different mother tongues. $G$ is markedly *less* sensitive
  than $S$ to small language communities, since each contributes only an
  infinitesimal $p_i^2$ term, a district can gain several tiny languages
  without moving $G$ noticeably (Grin & Fürst, 2022, Sec. 4.1).

> Grin, F., & Fürst, G. (2022). Measuring Linguistic Diversity: A
> Multi-level Metric. *Social Indicators Research*.
> <https://doi.org/10.1007/s11205-022-02934-5>
>
> Greenberg, J. H. (1956). The Measurement of Linguistic Diversity.
> *Language*, 32(1), 109–115. <https://www.jstor.org/stable/410659>



```{r calculate-entropy}
calculate_diversity <- function(data, level) {
  data |>
    filter(language_level == level, district_code != "000", !grepl("Others", language_name)) |>
    group_by(state_code, state_name, state_name_harmonized, district_code, language_name) |>
    summarise(speakers = sum(total_persons), .groups = "drop_last") |>
    mutate(proportion = speakers / sum(speakers)) |>
    summarise(
      n_languages = n(),
      # log2, not ln (Grin & Fürst 2022 use ln), to match the base of
      # 2^shannon_entropy below.
      shannon_entropy = -sum(proportion * log2(proportion), na.rm = TRUE),
      effective_languages = 2^shannon_entropy,
      greenberg_index = 1 - sum(proportion^2),
      .groups = "drop"
    ) |>
    inner_join(
      census_2011_pca |>
        transmute(
          state_code = sprintf("%02d", state_code),
          district_code = sprintf("%03d", district_code),
          district = name
        ),
      by = c("state_code", "district_code")
    )
}

diversity_l1 <- calculate_diversity(census_2011_mother_tongue, "L1")
diversity_l2 <- calculate_diversity(census_2011_mother_tongue, "L2")

cat("L1 (Scheduled Languages):\n")
summary(diversity_l1$effective_languages)

cat("\nL2 (Dialects):\n")
summary(diversity_l2$effective_languages)
```

We can compare Shannon entropy (via effective number of languages) and the Greenberg
Diversity Index:

```{r compare-entropy-gdi, fig.width=6, fig.height=6}
levels <- list(L1 = diversity_l1, L2 = diversity_l2)
for (level in names(levels)) {
  d <- levels[[level]]
  cat(
    level, "correlation (effective_languages vs greenberg_index):",
    cor(d$effective_languages, d$greenberg_index), "\n"
  )
}

ggplot(diversity_l1, aes(greenberg_index, effective_languages)) +
  geom_point(alpha = 0.5, color = "firebrick") +
  labs(
    title = "L1: entropy vs Greenberg index",
    x = "Greenberg Diversity Index",
    y = "Effective languages (Shannon)"
  ) +
  theme_census()
```

```{r prepare-maps}
map_l1 <- attach_geometry(diversity_l1, 2011, geography = "district")
map_l2 <- attach_geometry(diversity_l2, 2011, geography = "district")

c(l1 = sum(!sf::st_is_empty(map_l1$geometry)), l2 = sum(!sf::st_is_empty(map_l2$geometry)))
```

```{r plot-l1-diversity, fig.width=16, fig.height=12}
plot_map(
  map_l1,
  fill_var = "effective_languages",
  title = "Shannon: Scheduled languages (L1)",
  subtitle = "Effective number of languages per district",
  legend_title = "Effective\nLanguages",
  palette = "reds",
  show_state_boundaries = TRUE,
  trans = "log2",
  limits = c(1, 18),
  breaks = c(1, 2, 4, 8, 16)
) | plot_map(
  map_l1,
  fill_var = "greenberg_index",
  title = "Greenberg: Scheduled languages (L1)",
  subtitle = "Greenberg Diversity Index per district",
  legend_title = "Greenberg\nIndex",
  palette = "reds",
  show_state_boundaries = TRUE,
  limits = c(0, 1)
)
```

```{r plot-l1-count, fig.width=10, fig.height=12}
plot_map(
  map_l1,
  fill_var = "n_languages",
  title = "Number of scheduled languages spoken (L1)",
  subtitle = "Count of distinct scheduled languages per district",
  legend_title = "Languages",
  palette = "blues",
  show_state_boundaries = TRUE
)
```

```{r plot-l2-diversity, fig.width=16, fig.height=12}
plot_map(
  map_l2,
  fill_var = "effective_languages",
  title = "Shannon: Dialects (L2)",
  subtitle = "Effective number of dialects per district",
  legend_title = "Effective\nLanguages",
  palette = "reds",
  show_state_boundaries = TRUE,
  trans = "log2",
  limits = c(1, 35),
  breaks = c(1, 2, 4, 8, 16, 32)
) | plot_map(
  map_l2,
  fill_var = "greenberg_index",
  title = "Greenberg: Dialects (L2)",
  subtitle = "Greenberg Diversity Index per district",
  legend_title = "Greenberg\nIndex",
  palette = "reds",
  show_state_boundaries = TRUE,
  limits = c(0, 1)
)
```

```{r plot-l2-count, fig.width=10, fig.height=12}
plot_map(
  map_l2,
  fill_var = "n_languages",
  title = "Number of dialects spoken (L2)",
  subtitle = "Count of distinct dialects per district",
  legend_title = "Dialects",
  palette = "blues",
  show_state_boundaries = TRUE
)
```

```{r compare-levels}
cat("Top 10 by scheduled languages (L1):\n")
diversity_l1 |>
  arrange(desc(effective_languages)) |>
  select(state_name_harmonized, district, effective_languages) |>
  head(10)

cat("\nTop 10 by dialects (L2):\n")
diversity_l2 |>
  arrange(desc(effective_languages)) |>
  select(state_name_harmonized, district, effective_languages) |>
  head(10)
```

# Readings

For India specifically, Jolad & Agarwal document how the Census's own
categorisation (grouping "mother tongues" under scheduled languages) masks
much of this diversity:

> Jolad, S., & Agarwal, A. (2022, March 8). India's Linguistic Diversity:
> How the Census obscures it. *The India Forum*.
> <https://www.theindiaforum.in/article/what-census-obscures>
>
> Jolad, S., & Agarwal, A. (2021). Mapping India's Language and Mother
> Tongue Diversity and its Exclusion in the Indian Census.
> <https://doi.org/10.31235/osf.io/sjxc6>
